---
title: "RNA Seq analysis using clusterProfiler with gene ontology (GO)"
author: "Oscar Montoya"
date: "March 29, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


Libraries
```{r}

library(DOSE)
library(GO.db)
library(org.Hs.eg.db) # Annotation package
#source("https://bioconductor.org/biocLite.R")
#biocLite("topGO") 
library(topGO)
#biocLite("GSEABase")
library(GSEABase)
library(clusterProfiler)
#browseVignettes(package = "clusterProfiler")
```


An annotation package is required. I'll use "org.Hs.eg.db", specific for H. sapiens (taken from clusterProfiler vignette)
```{r}
#biocLite("org.Hs.eg.db")
library(org.Hs.eg.db)
browseVignettes(package = "org.Hs.eg.db")
?org.Hs.eg.db

keytypes(org.Hs.eg.db)

#http://bioconductor.org/packages/release/BiocViews.html#___OrgDb

```


Import expression results as a csv.
```{r}
pcks <- read.csv("/path/to/file.csv", header = TRUE, stringsAsFactors = FALSE)
str(pcks)
head(pcks)

#No duplicates gene symbols in pcks data set (three ways to test)
pcks$Symbol[duplicated(pcks$Symbol, incomparables = FALSE)]

which(pcks$Symbo[duplicated(pcks$Symbol, incomparables = FALSE)] == TRUE) 

pcks$Symbol[anyDuplicated(pcks$Symbol, incomparables = FALSE)]
```


If needed find ENTREZID for further data analyses and plots:

```{r}

get.ENTREZID = bitr(pcks$SYMBOL, fromType="SYMBOL", toType="ENTREZID", OrgDb="org.Hs.eg.db")

head(get.ENTREZID)
str(get.ENTREZID)

# Not all symbols are mapped. Find which symbols were not mapped:


setdiff(pcks$SYMBOL, get.ENTREZID$SYMBOL) #These are the symbols that weren't mapped


# Now I need to use merge() to fuse the two datasets while excluding the not mapped symbols.

pcks_1 <- merge(pcks, get.ENTREZID, by =
                  intersect(pcks$SYMBOL,
                            get.ENTREZID$SYMBOL), 
                by.x = "SYMBOL", 
                by.y = "SYMBOL")

head(pcks_1)

```

## Using Gene Ontology (GO)

### GO Classification

groupGO is designed for gene classification based on GO distribution at a specific level. Only expressions greater than 2 and lower than -2 will be analyzed ("expressions") and full dataset will be used for background genes -universe, in clusterProfiler ("all_genes") 


```{r}

# Take a look at the strcture used in clusterProfiler tutorial, called geneList from DOSE package

data(geneList, package="DOSE")
length(geneList) # 12495 genes in total

gene <- names(geneList)[abs(geneList) > 2]

length(gene) # 207 genes


?groupGO


# All names will be used in argument "universe"
all_genes <- pcks$Exp.Log.Ratio

names(all_genes) <-
  as.numeric(pcks$Entrez.Gene.ID.for.Human)


# "expressions_vector" vector will be used in argument "gene"
expressions <- all_genes[abs(all_genes) > 2] # 103 genes wih absolute expresions > 2

names(expressions)
length(expressions) 

# Test with geneList
test_geneList <- groupGO(gene = names(geneList)[abs(geneList) > 2],
               OrgDb    = org.Hs.eg.db,
               ont      = "CC",
               level    = 3,
               readable = TRUE)

head(test_geneList) 


# Actual analysis with my data

ggo_BP <- groupGO(gene = names(expressions),
               OrgDb    = org.Hs.eg.db,
               ont      = c("BP"), #Biological Process (BP), Cellular Component (CC) or molecular Function (MF) subontologies
               level    = 3,
               readable = TRUE)

head(ggo_BP); View(ggo_BP)

################

ggo_CC <- groupGO(gene = names(expressions),
               OrgDb = org.Hs.eg.db,
               ont = c("CC"), 
               level = 3,
               readable = TRUE)

head(ggo_CC); View(ggo_CC)

################

ggo_MF <- groupGO(gene = expressions,
               OrgDb = org.Hs.eg.db,
               ont = c("MF"), 
               level  = 3,
               readable = TRUE)

head(ggo_MF ); View(ggo_MF )

#################

head(ggo_BP); head(ggo_CC); head(ggo_MF)

```


## GO over-representation test


```{r}
?enrichGO

# Overrepresentation using biological processes (BP) ontology from Gene Ontology

keytypes(org.Hs.eg.db)


# ==============================

OverRep_GO <- enrichGO(gene = names(expressions),
                       universe = names(all_genes), 
                       OrgDb = org.Hs.eg.db,
                       #keytype = "ENTREZID",
                       ont = "BP",
                       pAdjustMethod = "BH",
                       pvalueCutoff = 0.5,
                       qvalueCutoff = 0.2 # falsexpressionse discovery rate (FDR) estimation from a collection of p-values or from a collection of test-statistics with corresponding empirical null statistics. From "qvalue" package.
                       #readable = TRUE
                       )
head(OverRep_GO)

# Test with geneList, as done in the tutorial

ego <- enrichGO(gene          = gene,
                universe      = names(geneList),
                OrgDb         = org.Hs.eg.db,
                ont           = "CC",
                pAdjustMethod = "BH",
                pvalueCutoff  = 0.01,
                qvalueCutoff  = 0.2,
                readable      = TRUE)
head(ego)

```


## GO Gene Set Enrichment Analysis (GSEA)

Not working yet. Not clear what to put as geneList

```{r}

?gseGO

#gswGO needs data sorted decreaing based on expresion levels



GSEA_GO <- gseGO(geneList = all_genes,
                       keytype = "ENTREZID",
                       OrgDb = org.Hs.eg.db,
                       ont = "CC",
                       nPerm = 1000,
                       #minGSSize = 10,
                       #maxGSSize = 500,
                       pvalueCutoff = 0.05, 
                 #by = "fgsea",
                       verbose = FALSE)

?gseGO
head(GSEA_GO)


```

## GO semantics similarity analysis
Useful to compare two sets of genes.

GO semantic similarity can be calculated by GOSemSim1. We can use it to cluster genes/proteins into different clusters based on their functional similarity and can also use it to measure the similarities among GO terms to reduce the redundancy of GO enrichment results.

http://bioconductor.org/packages/release/bioc/vignettes/GOSemSim/inst/doc/GOSemSim.html
```{r}
#source("https://bioconductor.org/biocLite.R")
#biocLite("AnnotationHub")
library(AnnotationHub)
hub <- AnnotationHub()

q <- query(hub, "Homo sapiens")
id <- q$ah_id[length(q)]
Hsapiens <- hub[[id]]

```

build annotation data needed
```{r}
library(GOSemSim)
HsapiensGO <- godata('org.Hs.eg.db', ont = "MF")
```

goSim and mgoSim... These two seem more two compare sets of genes instead of just looking at expression...?


# Using MSigDB gene set collections

```{r}
# From tutorial
#gmtfile <- system.file("extdata", "c5.cc.v5.0.entrez.gmt", package="clusterProfiler")

# Download from broad institute and read
c5 <- read.gmt("/path/to/genesets.gmt")

head(c5)
c5$gene


egmt <- enricher(names(expressions_symbol),
                 pvalueCutoff = 0.5,
                 pAdjustMethod = "BH",
                 #universe = names(all_genes_symbol),
                 minGSSize = 1, 
                 maxGSSize = 1000, 
                 qvalueCutoff = 0.5,
                 TERM2GENE=c5)

head(egmt)

################

egmt2 <- GSEA(all_genes_symbol,
              nPerm = 10000,
              pvalueCutoff = 0.9,
              pAdjustMethod = "BH",
              TERM2GENE=c5,
              #seed = 1,
              verbose=FALSE)

head(egmt2)
dim(egmt2)

enrichMap(egmt, 
          #n=15, 
          vertex.label.font = 0.1)

str(egmt)
head(egmt)
class(egmt)
reshape::melt(egmt) 
```



# Visualization

```{r}
?enrichMap
?cnetplot

str(OverRep_GO)

enrichMap(OverRep_GO)

# A less dense enrichMap plot by simpliying OverRep_GO. enrichMap uses netplot to buil the graph (from DOSE), which in turn uses iGraph.
simplified_OverRep_GO <- simplify(OverRep_GO, cutoff=0.9, by="p.adjust", select_fun=min)
enrichMap(simplified_OverRep_GO, vertex.label.font = 0.5)



# Facet_wrap the plot for easier visualization by directly using iGraph
library(igraph)
layouts <- grep("^layout_", ls("package:igraph"), value=TRUE)[-1] 

par(mfrow=c(3,3), mar=c(1,1,1,1))

for (layout in layouts) {

  print(layout)

  l <- do.call(layout, list(OverRep_GO)) 

  plot(OverRep_GO, edge.arrow.mode=0, layout=l, main=layout) }

```

## cnetplot of increase and decreased
```{r}
par(mfrow=c(1,1))

cnetplot_increased <- cnetplot(OverRep_GO, #categorySize="pvalue",
                               categorySize="geneNum",
                               foldChange=pcks_1$Exp.Log.Ratio[pcks_1$Exp.Log.Ratio>=2])

cnetplot_decreased <- cnetplot(OverRep_GO, categorySize="pvalue", foldChange=pcks_1$Exp.Log.Ratio[pcks_1$Exp.Log.Ratio<=-2])


plotGOgraph(OverRep_GO)

gseaplot(kk2, geneSetID = "hsa04145")

browseKEGG(kk, 'hsa04110')

library("pathview")
hsa04110 <- pathview(gene.data  = geneList,
                     pathway.id = "hsa04110",
                     species    = "hsa",
                     limit      = list(gene=max(abs(geneList)), cpd=1))

barplot(OverRep_GO, drop=TRUE, showCategory=12)

dotplot(OverRep_GO)

gseaplot?
```



